MLH1 Results Outline

### 2020-02-19 doc/Results.rmd
notes before knitting again

# Data set ups

@@

In order to estimate the genome wide recombination rates across our panel of three subspecies from the house mouse species complex, _M. m. domesticus,_ _M. m. musculus_. and _M. m. molossinus_. 
`r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild derived inbred strains were used with reflect samples of natural genetic variation across the terretories of these subspecies.

MLH1 foci were quantified from a total of `r sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "male"))$Ncells)` spermatocytes and `r sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells)` oocytes. The sampling of prophase oocytes is diffecult and was the limiting factor thus there are some strains with only MLH1 counts from spermatocytes, which are included for sex specific analysises.

1. -female strain means are v close  -male strain means are more varied!!
-Comparisons to references / literatures
-house mouse close to minimum of 1  obligate CO per chrm

2. The largest female difference is G, 1.07X  the largest male difference is PWD and MSM which are 1.3 and 1.2 respectively.

<!-- female -->
In order to more finely describe the sex specific patterns, we focused on data sets including single sexes.

@@

Within mouse variance was assess to follow up on the results from the full dataset.

@@

The variance across mouse means for each strain is also variable across strains. With _domesticus^LEW^_ having the most significant strain effect in the second model (p= ).

The overall pattern for the female mean MLH1 counts per cell is low variation. 
(the most significantly differnt strains in the female data), For _domesticus^G^_ the G strain is `r G.female.diff` higher the the other means. While _domesticu^LEW^_  is `r round(LEW.female.diff,4)` higher  and _musculus^MSM^_ is `r MSM.female.diff` higher compared to the other strain means.

 mouse mean ranges
In the _domesticus_ subspecies the ranges of mouse means are 
`r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "WSB"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "WSB"))$mean_co, na.rm=TRUE  )` in _domesticus^WSB^_ , `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "G"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "G"))$mean_co, na.rm=TRUE  )` 
in _domesticus^G^_ `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "LEW"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "LEW"))$mean_co, na.rm=TRUE  )` in _domesticus^LEW^_.

@@

While the male pattern shows more variation across the strain means.

@@

@@

The strain average for  _musculus^PWD^_ , _molossinus^MSM^_ and  _musculus^SKIVE^_ are  `r mean(PWD_male$mean_co, na.rm = TRUE) /mean(sans_LewRec_male$mean_co, na.rm = TRUE)`,

@@

<!--
$$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex + rand(strain) + \varepsilon $$

@@

```{r MLH1co.MM, echo=FALSE, warning = FALSE, digits = 3}
#THIS IS THE NEW MIXED MODEL
#should be able to use the loaded 
#DF.HetC.MixedModel.HQ

@@

All of the effects from the mixed model significant (Table X) so we followed up with general linear models. The first glm contained subspecies, strain, sex, and all interactions, while the second glm, contains coefficients for strain, sex and the interactions. (*what are the points and differences? )*

@@

@@

@@

@@

LmerMLH1_M1_Q12_VAR_results <- data.frame(sub.sp = LmerMLH1_M1_Q12_CV_subsp,
sex.results =  LmerMLH1_M1_Q12_CV_sex, interaction.pval = LmerMLH1_M1_Q12_CV_int,  rand.pval = LmerMLH1_M1_Q12_CV_strain)
```
<!--
$$mouse\ variance\ in \ CO \ per cell  ~=~ sex * strain + \varepsilon$$-->
```{r M3.MLH1.var, echo=FALSE}

@@

For both the mixed model and linear model, sex was the only significant effect while the subspecies, while interaction and random strain effects were not significant (Table X) (LRT; p= `r LmerMLH1_VAR_results$sex.results`, glm; p= `r MLH1var_M3_Q12_sum$coefficients[,4][2]` mixed and glm models respectively).

Since the measures for within mouse variance may be more susceptible to technical error effects from the staining protocol (i.e. background noise). For this reasons, we replicated the model analysis from data sets in the subset of cells with higher quality scores compared to the full data set. (these analysis/ models replicated the results of the full data set (sex being the most significant 
(LTR; p = `r LmerMLH1_M1_Q12_VAR_results$sex.results`, and glm; p = `r  MLH1var_M3_Q12_sum$coefficients[2,4]` )

@@

For male specific glms for within mouse variance -- had no significant strain effects

the female data models  --- inconsistant and moderate significance p values

@@

$$mouse\ variance\ for\ MLH1\ foci\ per\ cell   ~=~ subsp * sex + rand(strain)+ \varepsilon$$ -->
<!--

RW's main qtl paper used the within animal variance, 
"We mapped QTL for
MLH1 count variance within individuals, applying an approach
to deconvolve the mean-variance relationship (see
Materials and Methods)."

The fig5 also has a histogram of within animal variances in MLH1 counts (For F2s, with the parents as reference) (I could do this for males and females)
QTL on chrm 14, decreased variance in GG genotype, compared to hetZ and WSB (I think this also raised the mean)

@article{wang2017genetics,
  title={Genetics of genome-wide recombination rate evolution in mice from an isolated Island},
  author={Wang, Richard J and Payseur, Bret A},
  journal={Genetics},
  volume={206},
  number={4},
  pages={1841--1852},
  year={2017},
  publisher={Genetics Soc America}
}
references for phenotypic variance connected to robustness to environment, increase in developmental consistency
Waddington 1942, Rutherford and Lindquist 2998, Mackay and Lyman 2005, Levy and Siegal 2008)
phenotypic variance leading to adaptation - 
(Losick and desplan 2008, Beaumont 2009, Eldar and Eloqitz 2010)
Previous vQTLs (Shen etal 2012, Hulse and Cai 2013, Mulder et al 2016)
meiotic Checkpoints for ensuring consistency (Hochwagen and Amon 2006, Li and Schimenti 2007, Jaramillo-Lambert 2010 and Kauppi 2013)
-->

@@

In an attempt to locatize the above male specific crossoever number evolution within the meiotic pathway we quantified a marker for DSBs, DMC1, in early prophase meiocytes.

DMC1 foci were scored from a total of `r sum( (DMC1.mean.table %>%  filter(stage == "L") )$ncells)` leptotene and `r sum( (DMC1.mean.table %>%  filter(stage == "Z") )$ncells)` zygotene staged spermatocytes of (X for each strain) juvenile mice (12 to 18 days).

A subset of strains were choosen to test for differences between the high and low recombination groups; _musculus^PWD^_, _molossinus^MSM^_, _ musculus^KAZ^_ ,  _domesticus^WSB^_ , and _domesticus^G^_. One mouse was sampled for each strain.

As can be seen in the figure, the high recombining groups have significantly more DMC1 for for leptotene (t.test; `r ttest.HighLow.L.pval` , one-way-anova; `r round(l.anova.pval, 5)`), but not the zygotene cells (t.test; `r ttest.HighLow.Z.pval`, one-way-anova; `r z.anova.pval`).

The ratios for MLH1:DMC1^Leptotene^ and MLH1:DMC1^zygotene^ are shown in Table X, and can be used as a proxy for the CO:NCO ratio (instead CO:(CO+NCO) ratio). Notably there is not much variation and do not show a significant difference between the high and the low recombining groups (t.test; p = `r DMC1.Lratio_t.test$p.value` and `r DMC1.Zratio_t.test$p.value` for leptotene and zygotene based ratios respectively). These results indicate that the best predictor for CO number may be established before DSB formation (cite Baier et al 2014).

@@

#proportion_stats_category,  tot_biv, ncells

#add number of cells and bivalents in the corner

@@

sum( MELT_Chrm.Proportion$n1CO) / total.biv.all
sum( MELT_Chrm.Proportion$n2CO) / total.biv.all
```

@@

**most of the male proportions are 2/3's 1CO and 1/3 2COs. **

@@

**The high rec strains are distinguishable by having more 2COs at the cost of 1COs, with closer to 1:1**

@@

In order to better deconstruct the genome wide recombiation rate, we decided to deconstruct cell wide average by examining single chromosome patterns.

From our total set of cell images `r length(new.Curate.FULL.org$Obj.ID)` chromosome objects were isolated by the image analysis software. After the human curation step, `r length(new.Curate.FULL.org$Obj.ID)` single bivalent measures were left.

@@

@@

Table X illustrates the break down of single bivalent observations by category

Error compared to human measures and other details have been discribed in (Peterson_2019).

@@

The mouse averages for 3 bivalent level metrics will be analyzed across these questions:  
1)SC lengths (or total.SC)/chromatin compaction (using all cells regaurdless of CO number) 2) Interfocal Distance of 2CO bivalents and 3) the normalized foci position from 1CO bivalents.

@@

(We will first list the results for the Q1)

These are our predictions for the sex specifici patterns on the single bivalent lanscape

 Q1 Analysis, Predictions for Heterochiasmy Q1

Two bivalent level traits are predicted to display heterochiasmy (ie significant effects of sex);

1) SC length will be sexually dimorphic (sex effect will be significant)(cite Lynn)

2. Normalized 1CO positions will be sexually dimorphic (sedell and Kirkpatrick).

3) Interference / IFD will not be sexually dimorphic. Previous physical measures of interference were not different between sexes (deBoer et al 1996, petkov 2001).

@@

@@

@@

<!--Brief lit review for SC length and MLH1 expections.
add XX adjusment / arguments

In order to test one of the major/known correlates with genome wide recombiation rate variation, we compared the average

@@

@@

The most averages for total SC (skeletonizes) are significantly longer in females for all strains (p=
`r Q1_tot.sc_t.test.WSB$p.value`, `r Q1_tot.sc_t.test.G$p.value`, and `r Q1_tot.sc_t.test.LEW$p.value`
for the _mmdomesticues^WSB^ _, _mmdomesticues^G^ _, and _mmdomesticues^LEW^ _ respectively).

>>>>

# Data Set Up Notes

@@

In order to estimate the genome wide recombination rates across our panel of three subspecies from the house mouse species complex, _M. m. domesticus,_ _M. m. musculus_. and _M. m. molossinus_ using `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild derived inbred strains. MLH1 foci were quantified from a total of `r sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "male"))$Ncells)` spermatocytes and `r sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells)` oocytes. The sampling of prophase oocytes is diffecult and was the limiting factor thus there are some strains with only MLH1 counts from spermatocytes, which are included for sex specific analysises.

<!--
transition first aragraph to strain specific
Main points
1. -female strain means are v close  -male strain means are more varied!!
-Comparisons to references / literatures
-house mouse close to minimum of 1  obligate CO per chrm

2. The largest female difference is G, 1.07X  the largest male difference is PWD and MSM which are 1.3 and 1.2 respectively.  
-->

<!-- transition from first part into female -->

@@

<!--
The variance across mouse means for each strain is also variable across strains. With _domesticus^LEW^_ having the most significant strain effect in the second model (glm; p= ).
 -->

@@

@@

@@

Intial examinations of the sex specific data revealed that.. there is more variety in across the strain means in meals compared to females.

The overall pattern for the female mean MLH1 counts per cell is low variation. 
(the most significantly differnt strains in the female data), For _domesticus^G^_ the G strain is `r G.female.diff` higher the the other means. While _domesticu^LEW^_  is `r round(LEW.female.diff,4)` higher  and _musculus^MSM^_ is `r MSM.female.diff` higher compared to the other strain means in females.

while in the male data the strain average for  _musculus^PWD^_ , _molossinus^MSM^_ and  _musculus^SKIVE^_ are  `r mean(PWD_male$mean_co, na.rm = TRUE) /mean(sans_LewRec_male$mean_co, na.rm = TRUE)`,

In the _domesticus_ subspecies the ranges of mouse means are 
`r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "WSB"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "WSB"))$mean_co, na.rm=TRUE  )` in _domesticus^WSB^_ , `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "G"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "G"))$mean_co, na.rm=TRUE  )` 
in _domesticus^G^_ `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "LEW"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "LEW"))$mean_co, na.rm=TRUE  )` in _domesticus^LEW^_.

@@

@@

Check the DF of the number cited above

<!--

$$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex + rand(strain) + \varepsilon $$

@@

```{r MLH1co.lmer.MM, echo=FALSE, warning = FALSE, digits = 3}
#THIS IS THE NEW MIXED MODEL
#should be able to use the loaded 
#DF.HetC.MixedModel.HQ

@@

All of the effects from the mixed model significant (Table X) 
All of the effects within the mixed model - came out as significant -- (LRT: subsp p= `r MLH1.MMlmer_results$blah`
sex p = `r MLH1.MMlmer_results$blah`, subspecies * sex p = `r MLH1.MMlmer_results$blah`).
(random LTR; strain p= `r MLH1.MMlmer_results$blah`).

@@

Following up with linear models which tested for (subsecies M1), and strain effects alone (M3)
indecated ... identified several signifcant sex by strain interaction effects, (MSM and PWD) (glm; PWD*male p= and MSM-male p = )

@@

<!-- transition gwRR to within mouse variance -->

@@

RENAME THE MODELS M2 subsp * strain \* sex

@@

LmerMLH1_M1_Q12_VAR_results <- data.frame(sub.sp = LmerMLH1_M1_Q12_CV_subsp,
sex.results =  LmerMLH1_M1_Q12_CV_sex, interaction.pval = LmerMLH1_M1_Q12_CV_int,  rand.pval = LmerMLH1_M1_Q12_CV_strain)
```
<!--
$$mouse\ variance\ in \ CO \ per cell  ~=~ sex * strain + \varepsilon$$-->

```{r M3.MLH1.var, echo=FALSE}

@@

For both the mixed model and linear model, sex was the only significant effect while the subspecies, while interaction and random strain effects were not significant (Table X) (LRT; p= `r round(LmerMLH1_VAR_results$sex.results, 5)`, glm; p= `r round(MLH1var_M3_Q12_sum$coefficients[,4][2], 5)` mixed and glm models respectively).

Since the measures for within mouse variance may be more susceptible to technical error effects from the staining protocol (i.e. background noise). For this reasons, we replicated the model analysis from data sets in the subset of cells with higher quality scores compared to the full data set. (these analysis/ models replicated the results of the full data set (sex being the most significant 
(LTR; p = `r round(LmerMLH1_M1_Q12_VAR_results$sex.results, 5)`, and glm; p = `r  round(MLH1var_M3_Q12_sum$coefficients[2,4], 5)` )

@@

For male specific glms for within mouse variance -- had no significant strain effects
the female data models  --- inconsistant and moderate significance p values

@@

$$mouse\ variance\ for\ MLH1\ foci\ per\ cell   ~=~ subsp * sex + rand(strain)+ \varepsilon$$ -->
<!--
RW's main qtl paper used the within animal variance,

@article{wang2017genetics,
  title={Genetics of genome-wide recombination rate evolution in mice from an isolated Island},
  author={Wang, Richard J and Payseur, Bret A},
  journal={Genetics},
  volume={206},
  number={4},
  pages={1841--1852},
  year={2017},
  publisher={Genetics Soc America}
}
references for phenotypic variance connected to robustness to environment, increase in developmental consistency

Waddington 1942, Rutherford and Lindquist 2998, Mackay and Lyman 2005, Levy and Siegal 2008)
phenotypic variance leading to adaptation - 
(Losick and desplan 2008, Beaumont 2009, Eldar and Eloqitz 2010)
Previous vQTLs (Shen etal 2012, Hulse and Cai 2013, Mulder et al 2016)
meiotic Checkpoints for ensuring consistency (Hochwagen and Amon 2006, Li and Schimenti 2007, Jaramillo-Lambert 2010 and Kauppi 2013)
-->

@@

explain why male oer female measures were used, -- mention some measures were made -- but the number of cells was so too small.

In an attempt to locatize the above male specific crossoever number evolution within the meiotic pathway we quantified a marker for DSBs, DMC1, in early prophase spermtaocytes. DMC1 foci were scored from a total of `r sum( (DMC1.mean.table %>%  filter(stage == "L") )$ncells)` leptotene and `r sum( (DMC1.mean.table %>%  filter(stage == "Z") )$ncells)` zygotene staged spermatocytes of (X for each strain) juvenile mice (12 to 18 days) when the proportion of early prophase cells is higher (cite). A subset of strains were choosen to test for differences between the high and low recombination groups; _musculus^PWD^_, _molossinus^MSM^_, _ musculus^KAZ^_ ,  _domesticus^WSB^_ , and _domesticus^G^_. One mouse was sampled for each strain.

(the differences in mean DMC1 foci across the 2 time points confirms expected results. (t.test for pooled DMC1 foci) )

The high recombining groups have significantly more DMC1 for for leptotene (t.test; p=`r ttest.HighLow.L.pval` , one-way-anova; p=`r round(l.anova.pval, 5)`), but not the zygotene cells (t.test; p=`r ttest.HighLow.Z.pval`, one-way-anova; p=`r z.anova.pval`).

The ratios for MLH1:DMC1^Leptotene^ and MLH1:DMC1^zygotene^ are shown in Table X, and can be used as a proxy for the CO:NCO ratio (instead CO:(CO+NCO) ratio).
Notably these ratios do not show a significant difference between the high and the low recombining groups (t.test; p = `r DMC1.Lratio_t.test$p.value` and p=`r DMC1.Zratio_t.test$p.value` for leptotene and zygotene based ratios respectively). These results indicate that the best predictor for CO number may be established before DSB formation (cite Baier et al 2014).

@@

#proportion_stats_category,  tot_biv, ncells
#add number of cells and bivalents in the corner

@@

#sum( MELT_Chrm.Proportion$n1CO) / total.biv.all
#sum( MELT_Chrm.Proportion$n2CO) / total.biv.all
```

@@

**The vast majority of all bivalents in house mouse have one or two crossoveres (1CO an 2CO), even across sexes**

- fits with low rec rates observed in house mouse.

(rates of 0CO, 3CO and 4CO, these are marginal chromsomes classes)

**most of the male proportions are 2/3's 1CO and 1/3 2COs. but The high rec strains are distinguishable by having more 2COs at the cost of 1COs, with closer to 1:1**

**pvales for chi square test, are larger for PSD-MSM compared to low rec strains within subsp, PWD-KAZ and MSM-MOLF**

**these results confirm the intuitive conclusion/result that the rapid male specific evolution in gwRR proceeds though the chromosome level/though increasing the number of 2CO bivalents at the expense of 1CO bivalents**

@@

@@

In order to better deconstruct the genome wide recombiation rate, we decided to deconstruct cell wide average by examining features of the recombiantion landscape at the single bivalent level. recombination landscape refers to the i) total length of the SC-AE (potential area for COs), ii) the position of single crossovers along bivalents and iii) the spacing between multiple COs on the same bivalent.

From our total set of cell images `r length(new.Curate.FULL.org$Obj.ID)` chromosome objects were isolated by the image analysis software. After the human curation step, `r length(new.Curate.FULL.org$Obj.ID)` single bivalent measures were left. (make sure these are the hand.foci bivalents).

how do these agree with other patterns of MLH1

@@

#compare total boxes and rate passed
#Curated_BivData  -- pass rate?

@@

Table X illustrates the break down of single bivalent observations by category 
Error compared to human measures and other details have been discribed in (Peterson_2019).

@@

@@

@@

<!-- breif background on longer female SC  (mouse, human) Gruhn-->

@@

EXPLAIN WHY USING TOTAL SC AREA --

turn out comparing the means of single bivalents is complicated by the fact we can't identify any of the bivalents. -- so we use total sc

@@

@@

#Curated_BivData99 <- Curated_BivData[Curated_BivData$hand.foci.count < 5,]
#Curated_BivData99 <- Curated_BivData99[(!is.na(Curated_BivData99$sex)),]

#CLEAN UP THE FULL DATA SET

@@

The most averages for total SC (skeletonizes) are significantly longer in females for all strains (p=
`r format(Q1_tot.sc_t.test.WSB$p.value, scientific = F )`, `r format(Q1_tot.sc_t.test.G$p.value, scientific=F )`, and `r Q1_tot.sc_t.test.LEW$p.value`
for the _mmdomesticues^WSB^ _, _mmdomesticues^G^ _, and _mmdomesticues^LEW^ _ respectively).
